Blaming the Typeless: \\ Scalable, Human-Centric Python Localization

### 2018-08-23 paper/main.tex
more suggestions for the evaluation section

\begin{figure}
\begin{tikzpicture}
  \begin{axis}[
      xlabel=Clustering threshold parameter,
      ylabel=\# of naturally occuring program clusters,
      xmax=1.25,
      legend style={at={(0.5,0.25)},anchor=north}
    %   ytick distance=1000,
    %   title=Linkage: single
    ]
    \foreach \language in {ocaml, python}
    {
      \addplot+[mark=none] table[x=threshold, y=\language_cluster_counts_single, col sep=comma]{cluster_counts.csv};
    }
\legend{NATE OCaml data, Our Python data}
\end{axis}
\end{tikzpicture}
\caption{Number of clusters when data is clustered by single linkage.
  A linkage tree is built using agglomerative clustering, and the resulting tree is broken up
  into clusters by thresholding the inconsistency coefficient. The inconsistency coefficient
  of a link is its distance, z-scored against other nearby links. TODO: find a way to include
  the Rosetta Code baseline in this graph? better not to devote multiple graphs to this issue.
  LM: the y axis could be made into a percentage of the dataset size, but that might be
  weird?}
\label{fig-diversity}
\end{figure}

In the most similar prior work, Learning to Blame, each program in the dataset
was a student's attempt at one of 23 different homework problems. Thus the
dataset was comparatively homogenous, making it unclear whether the results
would generalize to a more diverse dataset.

In our PythonTutor data by comparision, users wrote whatever programs they
wanted, so we
hypothesized that the data were more heterogeneous. To confirm this
quantitatively, we used agglomerative clustering both on our
data and on the Learning to Blame data; whichever dataset naturally splits
into more clusters is more heterogeneous.

@@

To cluster programs, we transform the program ASTs into flat strings, and then
perform agglomerative clustering on them, with string edit distance as a distance
metric.

@@

To measure the distance between programs, we flattened their ASTs into strings
of tokens, and then computed the Levenshtein edit distance~\cite{levenshtein}.
The decision to flatten the tree rather than compute a distance directly on the
the ASTs is practical; the state of the art in tree edit distance has cubic
asymptotic complexity~\cite{PAWLIK2016157}, while the levenshtein distance can be
computed in time quadratic in the length of the strings~\cite{lev-quadratic}.
When we attempted a test run using tree edit distance as the metric, it did not
complete and we determined that it would not be practical to use.

The flattening process, which is based on \cite{dist-site}, is defined
recursively: the string for an AST
begins with a unique token corresponding to the AST node type, which is
followed by the concatenation of the transformations of the tree's subtrees.
The string ends with an ``end'' token to denote the end of the tree. This
preserves the tree structure of the AST.

@@

We performed agglomerative clustering on the programs~\cite{modern-clustering}.
This is a method of clustering in which every datapoint starts in its own
cluster, and the two closest clusters are merged every step of the algorithm,
until every datapoint is in the same cluster. This produces a binary tree, where
each internal node corresponds to a cluster formed by merging two other clusters,
and each leaf corresponds to an item in the dataset. We used single linkage,
which means that the distance between clusters of multiple items was considered
to be the minimum distance between their elements.

This alone produces a tree rather than a list of disjoint clusters, which makes
different clusterings difficult to compare. To produce a cluster count, we used
the SciPy function \lt{scipy.cluster.hierarchy.fcluster}, which forms disjoint
clusters by splitting the tree at internal nodes where the two joined clusters
are more distant than some threshold. We used the \lt{'inconsistent'} method with
depth set to 2, which specifies that the distances at each node should be
z-scored against the distances at other nodes at the same level in the tree, and
any nodes 1 or 2 levels below~\cite{inconsistent}. This adapts to differences in
scale of edit distance that might arise between python and OCaml due to
programming language expressivity or other factors, making cluster counts at each
threshold value comparable.

\figref{diversity} shows the results of clustering on our dataset and the NATE
dataset. For all values of the inconsistency threshold, the algorithm computed a
higher cluster count for the Python data, suggesting that it contains a more
diverse set of programs than the OCaml dataset. The similarity in the graphs
appears to confirm the comparibility of cluster counts between the two datasets.

>>>>

\begin{figure}
\begin{tikzpicture}
  \begin{axis}[
      xlabel=Clustering threshold parameter,
      ylabel=\# of naturally occuring program clusters,
      xmax=1.25,
      legend style={at={(0.5,0.25)},anchor=north}
    %   ytick distance=1000,
    %   title=Linkage: single
    ]
    \foreach \language in {ocaml, python}
    {
      \addplot+[mark=none] table[x=threshold, y=\language_cluster_counts_single, col sep=comma]{cluster_counts.csv};
    }
\legend{NATE OCaml data, Our Python data}
\end{axis}
\end{tikzpicture}
\caption{Number of natural program clusters in each dataset as a function
of clustering parameter (informally, how ``strict'' or ``tight'' the
clustering is). \fixme{Talk about how you sample to be fair to the smaller
datasets.} At all points our dataset demonstrates more diversity than 
that of previous work.} 
\label{fig-diversity}
\end{figure}

To demonstrate that our evaluation dataset is not only larger but also more
diverse than those used in previous work, we compare the diversity of
programs present here to those in a relevant balance. The Nate
algorithm~\cite{learning-to-blame} also provides type error localization
and also focuses on a machine-learning approach backed by slicing. However,
Nate targets strongly statically typed OCaml programs. Nate's evaluation
also focused on beginner-written programs: specifically, it considered
student submissions to 23 different university homework problems.  Thus the
Nate dataset is comparatively homogeneous, raising concerns about whether
evaluation results would generalize to more diverse settings.

In our PythonTutor dataset, by comparison, users were not constrained to
specific university assignments. We hypothesize that the evaluation data are
more heterogeneous. To assess this quantitatively, we used agglomerative
clustering to discover the number of ``natural'' program categories
present in both our dataset and also in the Nate dataset. Datasets with
more distinct program categories are more heterogeneous.

@@

% To cluster programs, we transform the program ASTs into flat strings, and
% then perform agglomerative clustering on them, with string edit distance
% as a distance metric.

@@

Many clustering algorithms depend on distance metrics. 
To measure the distance between programs, we flattened their ASTs into
strings of tokens, and then computed the Levenshtein edit
distance~\cite{levenshtein}. The decision to flatten the tree rather than
compute a distance directly on the ASTs relates to scalability for our
large datasets. The state of the art in tree edit distance has cubic
asymptotic complexity~\cite{PAWLIK2016157}, while the Levenshtein distance
can be computed in time quadratic in the length of the
strings~\cite{lev-quadratic}. We did implement and attempt a tree edit
distance metric, but it did not terminate for this analysis in a practical
amount of time.

The flattening process, which is based on standard approaches for
tree-structured data~\cite{dist-site}, is defined recursively: the string
for an AST begins with a unique token corresponding to the AST node type,
which is followed by the concatenation of the transformations of the tree's
subtrees.  The string ends with an ``end'' token to denote the end of the
tree. This embeds information related to the tree structure of the AST in
the flattened string.

@@

We performed agglomerative clustering on the 
datasets of flattened programs~\cite{modern-clustering}.
This is a method of clustering in which every datapoint starts in its own
cluster, and the two closest clusters are merged every step of the algorithm,
until every datapoint is in the same cluster. This produces a binary tree, where
each internal node corresponds to a cluster formed by merging two other clusters,
and each leaf corresponds to an item in the dataset. We used a single linkage
approach in which the distance between clusters of multiple items was considered
to be the minimum distance between their elements.

To produce a cluster count from the agglomerative tree, we followed the
standard approach of splitting the tree at internal nodes where the two
joined clusters are more distant than some threshold. To account for
potential programming languages differences between Python and OCaml (such
as expressivity), we z-score nodes against not only others at the same tree
depth, but also against against others one or two levels
below~\cite{inconsistent}. This makes cluster counts at each threshold
value comparable. Our implementation uses the standard
SciPy library (e.g., \lt{scipy.cluster.hierarchy.fcluster} with the 
\lt{inconsistent} method).

\fixme{We need to say something about how we controlled for number, lest
reviewers think we have more clusters simply because we have hundreds of
thousands of datapoints.}

\figref{diversity} shows the results of clustering on our dataset and the
Nate dataset. At a high level, the structural similarity in the graphs
confirms the comparability of cluster counts between the two datasets.  For
all values of the inconsistency threshold, there are more clusters in our
Python dataset than in the baseline OCaml dataset. This suggests that our
dataset contains a more diverse set of programs. Note that we are not
claiming any novelty or advances in clustering accuracy in this
determination (indeed, scalability concerns limited us to coarser
approaches); instead, our claim is that even with simple clustering,
it is clear that our dataset contains a greater diversity of programs than
were considered by previously-published evaluations.  
We view it as an advantage of our algorithm that it can apply to multiple
different program categories.
